---
title: "Reproducible Research - Peer Assessment 1"
author: "FF"
date: "Wednesday, August 13, 2014"
output: html_document
---

### Part 1: Raw data analysis

In a first step, the data are read and the number and distribution of observations is analyzed. 

```{r}
library(knitr)
opts_chunk$set(cach.path = 'cache/PA1_FF')
```



```{r}
activity = read.csv("data/activity.csv")
# determine number of records per date
t1 <- table(activity$date, activity$interval)
dim(t1); min(t1); max(t1)
```

We see that there is exactly one record for every 5-minute interval of every day(61 days, 288 intervals per day).

However, there are `r sum(is.na(activity$step))` missing values for the variable "step". 

For further analysis, I convert the data into a 2D matrix (rows represent days and columns represent intervals).
This easily gives the counts per day. 
```{r}
tx = tapply(activity$steps,
            list(activity$date, activity$interval),
            sum)
count.per.day = rowSums(tx, na.rm=T)
avg.per.interval = colMeans(tx, na.rm=T)
```

Here is a histogram for the counts per day: 
```{r Histogram_Initial}
hist(count.per.day,
     breaks=20,
     main="Histogram 1: Counts per day - with missing values")
```

Mean of steps per day is `r mean(count.per.day, na.rm=T)`
and the median is `r median(count.per.day,na.rm=T)`.

The average activity per interval is as follows:
```{r Activity_all_days}
plot(x=(0:287)/12,          # scale per hour 
     y=avg.per.interval,
     xlab="Interval (scale is hours)",
     type="l",
     main="Average per interval")
```

### Part 2: Imputing missing values
As seen above, there are `r sum(is.na(activity$step))` missing values for the variable "step".

We use the standard impute() function per interval (=column in our matrix), so the missing values are replaced by the interval means. 

```{r results='hide'}
library(Hmisc)
```
```{r}
tx.imputed = tx
for(i in 1:dim(tx)[2])
    tx.imputed[,i] = impute(tx[,i])

count.per.day.imputed = rowSums(tx.imputed)
```

The total number of steps before imputing is 
`r sum(tx,na.rm=T)`, the total number of steps after imputing has increased to `r sum(tx.imputed)`.

Here is a histogram for the counts per day: 
```{r Histogram_imputed}
hist(count.per.day.imputed,
     breaks=20,
     main="Counts per day - after imputing")
```

Mean of steps per day is now `r mean(count.per.day.imputed)`
and the median is `r median(count.per.day.imputed)`.

*Since the imputation only changed days with few steps, the median has stayed the same, but the mean has gone up.*

### Part 3: Weekdays vs Weekend Analysis 

For this part of the analysis, we perform the following steps:

* convert the matrix to a data frame

* operate on the row names (=dates) go get the weekdays

* create a logical variable indicating the weekends 

* compute the average steps per group 
 
```{r}
txd = as.data.frame(tx.imputed)
days = as.POSIXlt(strptime(rownames(txd),"%Y-%m-%d"))
weekend = days$wday >= 5     # use POSIX weekday numbers 

txd.s = split(txd, weekend )     # split weekday / weekend
names(txd.s) = c("weekday","weekend")  # assign names to split
timelines.s = sapply(txd.s, colMeans)  # compute avg. per split
```

Now we could simply plot the two patterns for the weekdays and the weekend. 

In order to get the required lattice graphic layout, we convert the data to a "long" data frame and we convert the time labels to fractions of hours. 
 
```{r}
df.long =  as.data.frame.table(timelines.s) # get long data frame 
# adjust names 
names(df.long)[1:3] = c("Interval", "Group", "Steps.avg") 
# compute fractions of hour
Int.num = as.numeric(as.character(df.long$Interval))
# separate into hours, minutes and combine 
df.long$Hour = floor(Int.num/100) + (Int.num %% 100)/60
```

Now we can create the lattice plot 
```{r Acitivity_weekday_vs_weekend}

xyplot(Steps.avg ~ Hour | Group,
       data=df.long,
       type="l",
       xlab="Interval",
       xlim=c(0,24),
       ylab="Average number of steps",
       layout=c(1,2))

```
 


